---
title: "PEB Quick Start Guide"
author: "Jason Thorpe [ jdthorpe_at_gmail_dot_com ]"
date: "`r Sys.Date()`"
output:
  html_document
vignette: >
  %\VignetteIndexEntry{PEB Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Calculating Required Parameters 

When using the Parametric Empirical Bayes (PEB) algorithm to set
personalized thresholds for a biomarker, it is necessary estimate the
population average ($\mu$,`mu`) for the biomarker as well as the within- and
between-person variation of that marker (sigma squared [$\sigma^2$,`s2`] and
tau squared[$\tau^2$,`t2`], respectively).  A dataset with observed marker
levels from a population of unaffected individuals is required, with multiple
observations for at least some observations.

The function `PEBparams()` calculates *intital* estimates for `mu`, `s2`,
and `t2`, as follows:


```{r,echo=FALSE}
#-- obs <- pmax(rpois(20,lambda=2.5),1)
#-- within <- rnorm(sum(obs)) 
#-- between <- rep(rnorm(20),obs)
#-- person = c(1,1,2,3,3,3,4,4,4,5)
#-- myData <- data.frame(person = factor(rep(letters[1:20],obs)),
#-- 					markerLevels = round(100 + within *3+ between*7,1))
#-- dput(myData)

myData <- structure(list(person = structure(c(1L, 1L, 2L, 2L, 2L, 3L, 3L, 
3L, 3L, 4L, 4L, 5L, 5L, 5L, 6L, 6L, 6L, 7L, 7L, 7L, 7L, 8L, 8L, 
8L, 8L, 8L, 8L, 9L, 9L, 10L, 11L, 11L, 12L, 12L, 13L, 13L, 13L, 
13L, 13L, 13L, 14L, 14L, 15L, 15L, 16L, 16L, 17L, 17L, 18L, 18L, 
18L, 18L, 19L, 19L, 20L, 20L, 20L), .Label = c("a", "b", "c", 
"d", "e", "f", "g", "h", "i", "j", "k", "l", "m", "n", "o", "p", 
"q", "r", "s", "t"), class = "factor"), markerLevels = c(104.6, 
98.8, 109.7, 107.7, 112.9, 102.7, 106.7, 110.4, 105.4, 93.8, 
91.3, 100.2, 105, 102, 100.5, 101.7, 103.6, 101.9, 102.3, 104.5, 
108.4, 100.6, 102.8, 97, 103.1, 100.1, 101.5, 102.6, 103.6, 104.2, 
99.4, 100.2, 101.2, 99, 104.8, 100.5, 104.9, 103.9, 102.4, 103.9, 
88, 87.8, 99.5, 91.7, 108.5, 103.1, 100.9, 104.3, 100.5, 95.5, 
96.4, 94.9, 99.2, 99.6, 93.9, 93.4, 94.8)), .Names = c("person", 
"markerLevels"), row.names = c(NA, -57L), class = "data.frame")
```

```{r}
library('PEB')
head(myData,10)
params <- PEBparams(x=myData$markerLevels,
					id=myData$person)
params
```

## Calculating a personalized threshold

In order to evaluate whether or not an individual's current marker levels
(`y_current`) is unexpectedly high, with fixed specificity given her personal
history ('y_past') and her membership in the population, a threshold can be
calculated using `threshold = qpeb(...)`, as follows:

```{r}
# let's use person 'a' as an example.  Her prior values are:
(y_past = myData$markerLevels [myData$person == 'a'])

# and a personalized threshold with 95% specificity can be set via:
qpeb(p=0.95,# specificity level
	 n=length(y_past),
	 ybar=mean(y_past),
	 mu=params$mu,
	 sigma2=params$s2,
	 tau2=params$t2)

# or simply:
qpeb(p=0.95,# specificity level
	 n=length(y_past),
	 ybar=mean(y_past),
	 params)
```

Hence, if her next marker level (`y_current`) exceeds 106.38, we can say that her marker is 
elevated with 95% specificity.

## Determining a personalized significance level

An alternative to calculating a threshold is to calculate the quantile
of the current value conditional on her history and the distribution of the
biomarker in the population using `ppeb(...)`, as follows:

```{r}
y_current <- 105.8
ppeb(y_current,
	 n=length(y_past),
	 ybar=mean(y_past),
	 params)
```

In the above calls to `qpeb()` and `ppeb()`, the number and average of the
previous biomarker level from the same individual were calculated explicitly
and passed to `ppeb()` as parameters.  However, if the parameters `n` and
`ybar` are not provided, `qpeb()` and `ppeb()` will calculate `n` and `ybar`
for each value in y using the assumption that y is a vector containing a
series of marker results from a **single** individual, as in:

```{r}
y <- c(y_past,
	   y_current)
ppeb(y,params)
```

## Accuracy of Specificity Levels

Because the accuracy the specificity for each threshold depends on the
sample sizes used to calculate the PEB parameters, it is useful to estimate
the accuracy of the specificity of PEB thresholds.  This can be accomplished by using the
'iterative' method to calculate the PEB parameters, and specifying the
`conf.level` parameter of `qpeb()`

```{r}
# calculate the parameters using the Iterative method
params_iterative <- 
	PEBparams(x=myData$markerLevels,
			  id=myData$person,
			  method='iterative')

qpeb(p = 0.95,
	params_iterative,
	n=length(y_past),
	ybar=mean(y_past),
	conf.level=0.9)
```

Notice that the threshold has changed slightly (106.44, up from 106.38)
with the 'iterative' method, and that the confidence interval for the false
positive rate is fairly wide (`[90.7%,98.6%]`), indicating that our initial
dataset with 57 observations from 20 individuals may be too small to
estimate thresholds with sufficient precision for many applications.

For a complete discussion on the PEB parameters, see `vignette('peb-parameters')`



